Dynamic Bandwidth Allocation with Minimum Rate Guarantees Using OpenFlow

### 2019-02-05 giformat/draft.tex
Added methodology intro, fixed some prev work stuff. Still need to explain technical limitations of pica8

\hspace{10mm} Software-Defined Networking (SDN) is an emerging network paradigm that decouples network control and forwarding functions, enabling network control to be directly programmable \cite{opennetworking}. This is done by implementing the network control function in a centralized controller node, which communicates to all the network devices in the network through a standardized protocol. OpenFlow is one such protocol; an open standard protocol that acts as an abstracted switch interface to allow the SDN controller to control the routers/switches in its domain \cite{sdx}.

\hspace{10mm} The purpose of this research is to design and implement an SDN solution to the excess bandwidth problem, in order to maximize the utilization of a network while providing minimum bandwidth guarantees to its clients. This is done by directing network traffic through an OpenFlow enabled switch. The switch is connected to a SDN controller device, which uses the switch's features to guarantee minimum bandwidths while dynamically allocating the excess bandwidth to clients according to a dynamic allocation algorithm, of which three have been developed. The algorithms in question are inspired by common CPU scheduling algorithms, but have been custom designed to approximate the functionality of their inspirations within the limitations of the OpenFlow switch environment.

\hspace{10mm} This research contributes to the body of QoS research by presenting an original SDN/OpenFlow traffic shaping design, which provides both minimum bandwidth as a guaranteed service, as well as the ability to utilize and fairly distribute the excess bandwidth in the link. Once implemented the solution will be evaluated first on its ability to guarantee minimum rates, and then on each algorithm's link capacity utilization and fairness of allocation.

@@

Designing methods of achieving hard bandwidth guarantees was the focus of research by Tomovic et al. \cite{Tomovic:2014}, Dwarakanathan et al. \cite{Dwara:2015}, and Egilmez et al. \cite{Egil}.
\\

\hspace{10mm} Tomovic et al. \cite{Tomovic:2014} argue that providing Quality of Service (QoS) guarantees to applications is an important objective of modern networks. With the rising popularity of real-time applications, especially in multimedia, the need arises for a network service model with performance guarantees. Current models (such as IntServ and DiffServ) lack adequate control, so QoS is currently often implemented by assigning separate physical networks with their own hardware to each class of traffic, incurring high operational expenditures and infrastructure costs. The paper aims to address this issue by designing a Software-Defined Networking (SDN) control environment to provide bandwidth guarantees between two points. The solution works by shifting network intelligence into a PoX SDN controller, which accepts/denies bandwidth requests, monitors network resources, and calculates traffic routes \cite{Tomovic:2014}. The solution is evaluated through a pair of experiments, which aim to demonstrate that the system can actually provide the minimum bandwidth guarantees between two points in the work. IntServ has been chosen as a baseline, and the results of each experiment are compared this baseline to measure improvement. This paper contributes to the body of QoS research by demonstrating the design of an SDN/OpenFlow control environment that provides bandwidth guarantees for priority flows in an automated manner, alleviating the need for distinct physical networks for multiple classes of service.

\hspace{10mm} Dwarakanathan et al. \cite{Dwara:2015} argue that current QoS solutions are not sufficiently scalable. To address this they implemented a similar SDN environment to that of Tomovic et al. \cite{Tomovic:2014}, using a Floodlight SDN controller as the point of network intelligence instead of a PoX controller. Dwarakanathan's \cite{Dwara:2015} solution aims to be more scalable than that of Tomovic et al. \cite{Tomovic:2014} by using fewer queues in each switch. Instead of creating one queue per guaranteed flow as in Tomovic's \cite{Tomovic:2014} system, it instead uses only one queue shared between all guaranteed flows, whose minimum and maximum are dynamically adjusted as bandwidth requests are granted and previously granted flows expire. The solution is evaluated through experiments ran in the Mininet simulator, in which theoretical minimum rates are calculated and compared against the experimental results. This paper contributes to the body of QoS research by demonstrating that it is possible to create a scalable QoS solution with SDN.

\hspace{10mm} In all of these systems, the maximum rate is equal to the guaranteed rate, preventing the flow from ever exceeding its guaranteed rate. This means that any unreserved bandwidth goes unused, resulting in sub-optimal link capacity utilization. They also both rely on all nodes in the network between the start and end points allocating resources, which introduces scalability difficulties.

@@

Designing methods of achieving soft bandwidth guarantees was the focus of research by Krishna et al. \cite{Krishna:2016}, Wallner and Cannistra \cite{Wallner:2013}, and Lee et al. \cite{SWLee}.
\\

\hspace{10mm} Krishna et al. \cite{Krishna:2016} argue that Quality of Service (QoS) is currently underutilized in provider networks, because it is too complicated to install and maintain. A response to this issue would be of great interest to large-scale internet service providers, who currently massively overprovision their networks at great cost. This paper aims to demonstrate how to use Software-Defined Networking (SDN) and the OpenFlow protocol’s QoS features to cheaply and efficiently implement minimum bandwidth guarantees in a network. The solution described in the paper uses a Ryu SDN controller communicating with an OpenFlow enabled switch, through which all QoS traffic will pass \cite{Krishna:2016}. The minimum bandwidth guarantee is applied by using a combination of the switch’s ingress meter table, and and egress queues. The ingress meters are used to aggregate traffic and redirect packets over the guaranteed minimum rate, and the egress queues and used to ensure priority traffic is transmitted first. The success of their solution is measured by the results of their three experiments. The first experiment tests the creation of new QoS flows in the system, the second tests traffic aggregation and prioritization, and the the third tests the responsiveness of the system. The experiments all have calculated theoretical expected results, and the effectiveness of the system is measured by how close the actual results come the expected results. The key metrics analyzed are bandwidth use for each flow, bandwidth use of the full channel, and time to adjust to changes in bandwidth usage. The paper contributes to the body of QoS knowledge by demonstrating how to implement minimum bandwidth guarantees using OpenFlow, while still allowing excess bandwidth to be shared without causing packet loss.

\hspace{10mm} Wallner and Cannistra \cite{Wallner:2013} argue that different network services (such as Voice Over IP (VOIP), stored video streaming, and live video streaming) have different requirements in terms of latency and bandwidth, and that these services can be divided into classes based on these requirements. Their paper addresses this by presenting a proof of concept model for class-based QoS control using SDN. A floodlight SDN controller is used to control many switches in a network, and in each switch incoming packets are classified based on their Type of Service (ToS) bits in the IP header. From this point different classes are given different priorities as they leave the switch. In this system minimum rates are only guaranteed by class and not by individual client, and the excess bandwidth is only distributed on a first-come first-serve basis. The effectiveness of this solution is unclear, as no empirical methods of evaluation are mentioned in the paper. The paper contributes to the body of QoS research by demonstrating that it is possible to use SDN to implement class-based QoS.

\hspace{10mm} In Wallner and Cannistra's \cite{Wallner:2013} system minimum rates are only guaranteed by class and not by individual client, and in both systems the excess bandwidth is only distributed on a first-come first-serve basis, allowing clients with high traffic to unfairly dominate the channel.

@@

Designing methods of performing real-time monitoring of flow statistics was the focus of papers by van Adrichem et al. \cite{vanAdrichem}, Ren and Xu \cite{RenOF14}, and Wang \cite{WangMon}.
\\

\hspace{10mm} Van Adrichem et al. \cite{vanAdrichem} argue that a key requirement to implement effective QoS in an SDN network is accurate traffic monitoring, on a finely-grained (per flow) basis. They state that current flow-based measurement techniques are too resource hungry in terms of CPU and bandwidth usage. To address this issue they propose \textit{OpenNetMon}, a POX OpenFlow controller module enabling accurate monitoring of per-flow throughput, packet loss and delay metrics \cite{vanAdrichem}. \textit{OpenNetMon} improves upon current measurement techniques by optimizing the choice of which switches are polled, and the frequency which they are polled, in order to minimize network resource usage while maintaining high accuracy \cite{vanAdrichem}. \textit{OpenNetMon}'s effectiveness was evaluated by running an \textit{OpenNetMon} module as well as \textit{tcpstat} on a node, and comparing their monitoring statistics as well as CPU and bandwidth usage. This paper contributes to the body of SDN research by demonstrating that is is possible to efficiently perform per-flow monitoring.

@@

Design, and evaluation of network scheduling algorithms was the focus of papers by Anderson et al. \cite{Anderson2004FairSO}, Al-Somaidai et al. \cite{Basheer}, Balogh et al. \cite{Balogh}, Swetha et al. \cite{Swetha}, and Borve \cite{Brve2009PacketSA}.
\\

\hspace{10mm} Balogh et al. \cite{Balogh} argue that in a networking ecosystem that is increasingly making use of class-based priority QoS, the way how waiting packets in network nodes are selected for output is a significant factor for providing that QoS. In their paper they investigate three priority queuing based scheduling algorithms. The algorithms are Priority Queuing (PQ), Weighted Round Robin with Priority Queueing (WRRPQ), and Low Latency Queuing (LLQ) \cite{Balogh}. These algorithms all support strict priority queueing, in which time critical priority data can be transmit first. PQ is a simple strict priority algorithm, in which priority packets are sent first, and all other packets are sent in first-come first-serve \cite{Balogh}. WRRPQ is a combination of PQ and Weighted Round Robin, in which priority packets are sent first, and other packets are sent according to Weighted Round Robin \cite{Balogh}. LLQ is an algorithm which considers the time required to send each packet to minimize latency \cite{Balogh}. The algorithms were evaluated by testing each one in a simulated network using different kinds of traffic. The results found that effectiveness of each algorithm depended on the kind of traffic (LLQ better for VOIP, WRRPQ better for fixed size packets). This paper contributes to the body of QoS research by showing that there are different optimal priority queuing algorithms for class-based QoS depending on the type of traffic.
\\\\

@@

@@

@@

>>>>

Software-Defined Networking (SDN) is an emerging network paradigm that decouples network control and forwarding functions, enabling network control to be directly programmable \cite{opennetworking}. This is done by implementing the network control function in a centralized controller node, which communicates to all the network devices in the network through a standardized protocol. OpenFlow is one such protocol; an open standard protocol that acts as an abstracted switch interface to allow the SDN controller to control the routers/switches in its domain \cite{sdx}.

The purpose of this research is to design and implement an SDN solution to the excess bandwidth problem, in order to maximize the utilization of a network while providing minimum bandwidth guarantees to its clients. This is done by directing network traffic through an OpenFlow enabled switch. The switch is connected to a SDN controller device, which uses the switch's features to guarantee minimum bandwidths while dynamically allocating the excess bandwidth to clients according to a dynamic allocation algorithm, of which three have been developed. The algorithms in question are inspired by common CPU scheduling algorithms, but have been custom designed to approximate the functionality of their inspirations within the limitations of the OpenFlow switch environment.

This research contributes to the body of QoS research by presenting an original SDN/OpenFlow traffic shaping design, which provides both minimum bandwidth as a guaranteed service, as well as the ability to utilize and fairly distribute the excess bandwidth in the link. Once implemented the solution will be evaluated first on its ability to guarantee minimum rates, and then on each algorithm's link capacity utilization and fairness of allocation.

@@

Designing methods of achieving hard bandwidth guarantees was the focus of research by Tomovic et al. \cite{Tomovic:2014} and Dwarakanathan et al. \cite{Dwara:2015}.
\\

Tomovic et al. \cite{Tomovic:2014} argue that providing QoS guarantees to applications is an important objective of modern networks. With the rising popularity of real-time applications, especially in multimedia, the need arises for a network service model with performance guarantees. Current models (such as IntServ and DiffServ) lack adequate control, so QoS is currently often implemented by assigning separate physical networks with their own hardware to each class of traffic, incurring high operational expenditures and infrastructure costs. The paper aims to address this issue by designing a Software-Defined Networking (SDN) control environment to provide bandwidth guarantees between two points. The solution works by shifting network intelligence into a PoX SDN controller, which accepts/denies bandwidth requests, monitors network resources, and calculates traffic routes \cite{Tomovic:2014}. The solution is evaluated through a pair of experiments, which aim to demonstrate that the system can actually provide the minimum bandwidth guarantees between two points in the work. IntServ has been chosen as a baseline, and the results of each experiment are compared this baseline to measure improvement. This paper contributes to the body of QoS research by demonstrating the design of an SDN/OpenFlow control environment that provides bandwidth guarantees for priority flows in an automated manner, alleviating the need for distinct physical networks for multiple classes of service.

Dwarakanathan et al. \cite{Dwara:2015} argue that current QoS solutions are not sufficiently scalable. To address this they implemented a similar SDN environment to that of Tomovic et al. \cite{Tomovic:2014}, using a Floodlight SDN controller as the point of network intelligence instead of a PoX controller. Dwarakanathan's \cite{Dwara:2015} solution aims to be more scalable than that of Tomovic et al. \cite{Tomovic:2014} by using fewer queues in each switch. Instead of creating one queue per guaranteed flow as in Tomovic's \cite{Tomovic:2014} system, it instead uses only one queue shared between all guaranteed flows, whose minimum and maximum are dynamically adjusted as bandwidth requests are granted and previously granted flows expire. The solution is evaluated through experiments ran in the Mininet simulator, in which theoretical minimum rates are calculated and compared against the experimental results. This paper contributes to the body of QoS research by demonstrating that it is possible to create a scalable QoS solution with SDN.

In all of these systems, the maximum rate is equal to the guaranteed rate, preventing the flow from ever exceeding its guaranteed rate. This means that any unreserved bandwidth goes unused, resulting in sub-optimal link capacity utilization. They also both rely on all nodes in the network between the start and end points allocating resources, which introduces scalability difficulties.

@@

Designing methods of achieving soft bandwidth guarantees was the focus of research by Krishna et al. \cite{Krishna:2016} and Wallner and Cannistra \cite{Wallner:2013}.
\\

Krishna et al. \cite{Krishna:2016} argue that QoS is currently underutilized in provider networks, because it is too complicated to install and maintain. A response to this issue would be of great interest to large-scale internet service providers, who currently massively overprovision their networks at great cost. This paper aims to demonstrate how to use Software-Defined Networking (SDN) and the OpenFlow protocol’s QoS features to cheaply and efficiently implement minimum bandwidth guarantees in a network. The solution described in the paper uses a Ryu SDN controller communicating with an OpenFlow enabled switch, through which all QoS traffic will pass \cite{Krishna:2016}. The minimum bandwidth guarantee is applied by using a combination of the switch’s ingress meter table, and and egress queues. The ingress meters are used to aggregate traffic and redirect packets over the guaranteed minimum rate, and the egress queues and used to ensure priority traffic is transmitted first. The success of their solution is measured by the results of their three experiments. The first experiment tests the creation of new QoS flows in the system, the second tests traffic aggregation and prioritization, and the the third tests the responsiveness of the system. The experiments all have calculated theoretical expected results, and the effectiveness of the system is measured by how close the actual results come the expected results. The key metrics analyzed are bandwidth use for each flow, bandwidth use of the full channel, and time to adjust to changes in bandwidth usage. The paper contributes to the body of QoS knowledge by demonstrating how to implement minimum bandwidth guarantees using OpenFlow, while still allowing excess bandwidth to be shared without causing packet loss.

Wallner and Cannistra \cite{Wallner:2013} argue that different network services (such as Voice Over IP (VOIP), stored video streaming, and live video streaming) have different requirements in terms of latency and bandwidth, and that these services can be divided into classes based on these requirements. Their paper addresses this by presenting a proof of concept model for class-based QoS control using SDN. A floodlight SDN controller is used to control many switches in a network, and in each switch incoming packets are classified based on their Type of Service (ToS) bits in the IP header. From this point different classes are given different priorities as they leave the switch. In this system minimum rates are only guaranteed by class and not by individual client, and the excess bandwidth is only distributed on a first-come first-serve basis. The effectiveness of this solution is unclear, as no empirical methods of evaluation are mentioned in the paper. The paper contributes to the body of QoS research by demonstrating that it is possible to use SDN to implement class-based QoS.

In Wallner and Cannistra's \cite{Wallner:2013} system minimum rates are only guaranteed by class and not by individual client, and in both systems the excess bandwidth is only distributed on a first-come first-serve basis, allowing clients with high traffic to unfairly dominate the channel.

@@

Designing methods of performing real-time monitoring of flow statistics was the focus of a paper by van Adrichem et al. \cite{vanAdrichem}.
\\

Van Adrichem et al. \cite{vanAdrichem} argue that a key requirement to implement effective QoS in an SDN network is accurate traffic monitoring, on a finely-grained (per flow) basis. They state that current flow-based measurement techniques are too resource hungry in terms of CPU and bandwidth usage. To address this issue they propose \textit{OpenNetMon}, a POX OpenFlow controller module enabling accurate monitoring of per-flow throughput, packet loss and delay metrics \cite{vanAdrichem}. \textit{OpenNetMon} improves upon current measurement techniques by optimizing the choice of which switches are polled, and the frequency which they are polled, in order to minimize network resource usage while maintaining high accuracy \cite{vanAdrichem}. \textit{OpenNetMon}'s effectiveness was evaluated by running an \textit{OpenNetMon} module as well as \textit{tcpstat} on a node, and comparing their monitoring statistics as well as CPU and bandwidth usage. This paper contributes to the body of SDN research by demonstrating that is is possible to efficiently perform per-flow monitoring.

@@

Design, and evaluation of network scheduling algorithms was the focus of a paper by Balogh et al. \cite{Balogh}.
\\

Balogh et al. \cite{Balogh} argue that in a networking ecosystem that is increasingly making use of class-based priority QoS, the way how waiting packets in network nodes are selected for output is a significant factor for providing that QoS. In their paper they investigate three priority queuing based scheduling algorithms. The algorithms are Priority Queuing (PQ), Weighted Round Robin with Priority Queueing (WRRPQ), and Low Latency Queuing (LLQ) \cite{Balogh}. These algorithms all support strict priority queueing, in which time critical priority data can be transmit first. PQ is a simple strict priority algorithm, in which priority packets are sent first, and all other packets are sent in first-come first-serve \cite{Balogh}. WRRPQ is a combination of PQ and Weighted Round Robin, in which priority packets are sent first, and other packets are sent according to Weighted Round Robin \cite{Balogh}. LLQ is an algorithm which considers the time required to send each packet to minimize latency \cite{Balogh}. The algorithms were evaluated by testing each one in a simulated network using different kinds of traffic. The results found that effectiveness of each algorithm depended on the kind of traffic (LLQ better for VOIP, WRRPQ better for fixed size packets). This paper contributes to the body of QoS research by showing that there are different optimal priority queuing algorithms for class-based QoS depending on the type of traffic.
\\\\

@@

This project implements a hybrid soft/hard QoS system, in a two-part solution. The first part of the solution is to achieve a minimum bandwidth guarantee for each client while retaining the use of the excess bandwidth on a first-come first-serve basis. The second part of the solution addresses the dynamic fair allocation of the excess bandwidth beyond only first-come first-serve. Both parts of the solution are implemented using an OpenFlow enabled Pica8 switch, with a Ryu SDN controller managing the details of the switch's flows, meters, and queues.

@@

put diagram here

@@

put other diagram here and put the tables back in
